The interactions of rational, pragmatic agents  \n lead to efficient language structure and use

### 2019-11-26 paper/paper.Rmd
last push before port

>>>>

We assume that speaker effort is related to the surprisal of an utterance in a particular context -- intuitively, the number of bits needed to encode the utterance $u$.\footnote{In the current set of simulations we consider utterances costs as independent from context (i.e. $p(u|c)=p(u)$). Hence surprisal can be thought of more simply as cost.} This particular formalization of speaker-cost is general enough to accommodate a range of cost instantiations, such as production difficulty via articulation effort, cognitive effort related to lexical access, or others (Bennett \& Goodman, 2015).\par

We assume listener effort is the surprisal of a meaning given an utterance. This operationalization of listener effort is intuitively related to existing work in sentence processing in which word comprehension difficulty is proportional to surprisal (Hale, 2001; Levy, 2008).

Rewriting (1) we have
\begin{equation}
\begin{split}
\text{Efficiency}(S, L, C, \ell) = \mathbb{E}_{e \sim P(E)}[-log_2(p(u|c))] + 
\mathbb{E}_{e \sim P(E)}[-log_2(L(m|u, c; \ell))]
\end{split}
\end{equation}

We assume that the particular joint distribution over utterance-meaning-context triples $e = <u, m, c>$  follows from a simple generative model: First, some context is sampled with probability $p(c)$. Then some meaning is sampled with probability $p(m|c)$. Our speaker attempts to convey this intended meaning to a listener via an utterance $u$ by sampling from the speaker conditional distribution $S(u|m, c; \ell)$. This allows us to re-write the objective as:

\begin{equation}
\begin{split}
  = -\sum_{u, m, c}P_{speaker}(u, m | c; \ell)p(c)[log_2(p(u|c))] - \\ 
   \sum_{u, m, c}P_{speaker}(u, m | c; \ell)p(c)[log_2(L(m|u, c; \ell))]
\end{split}
\end{equation}

\begin{equation}
\begin{split}
  = -\sum_{c\in C}p(c)(\sum_{u, m}P_{speaker}(u, m | c; \ell)[log_2(p(u|c))] + 
  \sum_{u, m}P_{speaker}(u, m | c; \ell)[log_2(L(m|u, c; \ell))])
\end{split}
\end{equation}

\begin{equation}
\begin{split}
 = -\sum_{c\in C}p(c)\sum_{u, m}P_{speaker}(u, m | c; \ell)[log_2(L(m|u, c; \ell)p(u|c)]
 \end{split}
\end{equation}

Note that $P_{listener}(u, m | c; \ell) = L(m|u, c; \ell)p(u|c)$:

\begin{equation}
\begin{split}
 = -\sum_{c\in C}p(c)\sum_{u, m}P_{speaker}(u, m | c; \ell)[log_2(P_{listener}(u, m | c; \ell))]
 \end{split}
\end{equation}

The inner summation of (6) is the cross-entropy between speaker and listener conditional distributions over utterance-meaning pairs.

\begin{equation}
\begin{split}
 = -\sum_{c\in C}p(c) H_{cross}(P_{speaker}(u, m |c; \ell), P_{listener}(u, m |c; \ell))
 \end{split}
\end{equation}

This final form is simply an expectation of speaker-listener cross-entropy over contextualized language use.

\begin{equation}
\begin{split}
 = \mathbb{E}_{c \sim P(C)}[H_{cross}(P_{speaker}, P_{listener}|c; \ell)]
 \end{split}
\end{equation}

Note that in the case that $|C| =1$, our objective simplifies to a simple Cross-Entropy between speaker-listener joint distributions over utterance-meaning pairs.\par

\begin{equation}
\begin{split}
 = H_{cross}(P_{speaker}, P_{listener}; \ell)
 \end{split}
\end{equation}

From an information-theoretic perspective this objective is intuitive: $H_{cross}$ denotes the Cross-Entropy (CE), a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.\par
